### 2025-07-15 sections/1-introduction.tex
intro edits

Researchers are collecting data at an unprecedented rate.
In many fields, datasets are growing exponentially, and this increase in the rate of data collection outpaces improvements in computing performance as predicted by Moore's Law~\cite{kahn2011future}.
This indicates that the performance of computing systems will not keep pace with the growth of data.
Often dubbed ``the Big Data explosion,'' this phenomenon has created a need for better algorithms to analyze large datasets.

Examples of large datasets include genomic databases, time-series data such as radio frequency signals, and neural network embeddings.
Large language models such as GPT~\cite{2020arXiv200514165B, OpenAI2023GPT4TR} and LLAMA-2~\cite{Touvron2023Llama2O}, and image embedding models~\cite{radford2021learning, dosovitskiy2020image} are a common source of neural network embeddings.
Among biological datasets,  SILVA 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 million genomes with an aligned length of 50,000 letters.
Among time-series datasets, the RadioML dataset~\cite{oshea2018radioml} contains approximately 2.55 million samples of synthetically-generated signals of different modulation modes.

Many researchers are interested in similarity search on these datasets.
Similarity search enables a variety of applications, including recommendation~\cite{annoy} and classification systems~\cite{suyanto2022knnclassifier}.
As the cardinalities and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become challenging;
even state-of-the-art algorithms exhibit a steep tradeoff between recall and throughput~\cite{malkov2016hnsw, johnson2019billion, annoy, aumuller2020ann}.

Given some measure of similarity between data points, there are two common definitions of similarity search: $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN).
$k$-NN search aims to find the $k$ most similar points to a query, while $\rho$-NN search aims to find all points within a similarity threshold $\rho$ of a query.
Previous works have used the term \textit{approximate} search to refer to $\rho$-NN search, but in this paper, we reserve the term \textit{approximate} for algorithms that do not exhibit perfect recall.
In contrast, an \textit{exact} search algorithm exhibits perfect recall.

$k$-NN search is one of the most ubiquitous classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}.
Na\"{i}ve implementations of $k$-NN search, whose time complexity is linear in the dataset's cardinality, prove prohibitively slow for large datasets.
While fast algorithms for $k$-NN search on large datasets exist, they are often approximate~\cite{gao2023high}, and while approximate search may be sufficient for some applications, the need for efficient and \textit{exact} search remains~\cite{ukey2023survey}.
For example, for a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but may be sensitive to local perturbations for smaller values of $k$.
This is especially true when classes are not well-separated~\cite{zhang2022imbalanced}.
Further, there is evidence that distance functions that do not obey the triangle inequality, such as cosine distance, perform poorly for $k$-NN search in biomedical settings~\cite{hu2016distance};
this suggests that approximate $k$-NN search could perform poorly in such contexts.

This paper introduces CAKES (CLAM-Accelerated $K$-NN Entropy-Scaling Search), a set of three novel algorithms for exact $k$-NN search.
We also compare CAKES against several current algorithms; namely, FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}, on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We further benchmark CAKES on a large genomic dataset, the SILVA 18S dataset~\cite{10.1093/nar/gks1219}, using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and a radio frequency dataset, RadioML~\cite{oshea2018radioml}, using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.
Finally, we compare real-world datasets to an artificially-generated dataset obeying simple statistical properties.

@@

Recent $k$-nearest neighbor search algorithms designed to scale with the exponential growth of data include Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, random projection and tree building (ANNOY)~\cite{annoy}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. However, some of these algorithms do not provide exact search (as defined in Section \ref{sec:introduction} above).

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} relies on navigable small world (NSW) networks~\cite{kleinberg2000navigation, boguna2009navigability} and skip lists.
HNSW builds a multi-layered graph of the dataset.
A query point and each data point are inserted into the graph and joined by an edge to the $M$ nearest nodes in the in the graph, where $M$ is a tunable parameter.
The highest layer in which an element can be placed is determined randomly with an exponentially-decaying probability distribution.
Search starts at the highest layer and descends to the lowest layer, greedily following a path of edges to the nearest node, until reaching the query point.
To improve accuracy, the $efSearch$ hyperparameter can be changed to specify the number of closest nearest neighbors to the query vector to be found at each layer.

InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} clusters data into high-dimensional Voronoi cells, and whichever cell a query point falls into is then searched exhaustively, similarly to an early precursor to our work~\cite{yu2015entropy}.
The number of cells used is governed by the $n_{list}$ parameter.
Increasing this parameter decreases the number of points being exhaustively searched, so it improves speed at the cost of accuracy.
To mitigate accuracy issues caused by a query point falling near a cell boundary, the algorithm has a tunable parameter $n_{probe}$ that specifies the number of additional adjacent or nearby cells to search.

ANNOY~\cite{annoy} relies on random projection and tree building for approximate $k$-NN search.
At each intermediate node of the tree, two points are randomly sampled from the space, and the hyperplane equidistant from them is chosen to divide the space into two subspaces.
This process iterates to create a forest of trees, and the number of iterations is a tunable parameter.
At search time, one can increase the number of trees to be searched to improve recall at the cost of speed.

@@

The entropy-scaling search paradigm exploits the geometric and topological structure inherent in large datasets.
Importantly, as suggested by their name, entropy-scaling search algorithms exhibit asymptotic complexity that scales with topological properties of the dataset, instead of its cardinality, when the dataset has a manifold structure.
In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search)~\cite{ishaq2019clustered}, which extended entropy-scaling $\rho$-NN search from a flat clustering approach to a hierarchical clustering approach.
CLAM (Clustering, Learning and Approximation with Manifolds), originally developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, is a refinement of the clustering algorithm from CHESS.
In this paper, we introduce CAKES, a set of three entropy-scaling algorithms for $k$-NN search, implemented in the Rust programming language.

We also provide a theoretical analysis of the time complexity of CAKES's algorithms in Sections~\ref{sec:methods:knn-search:repeated-rnn-complexity} and~\ref{sec:methods:knn-search:complexity-of-sieve-methods}. 
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a uniform distribution.
Given that CAKES's algorithms are intended to be used on datasets with a manifold structure, complexity analysis assuming a uniform distribution of data would be uninformative. 
As a result, we assume in our analyses that the dataset has a manifold structure, as quantified by low fractal dimension and metric entropy.
For our purposes, it makes more sense to analyze the expected performance of CAKES's algorithms under these assumptions.

>>>>

Researchers are collecting data at an unprecedented rate, with datasets in many fields growing exponentially, and outpacing improvements in computing performance predicted by Moore's Law~\cite{kahn2011future}. This ``Big Data explosion'' has created a need for algorithms that scale efficiently to large datasets.

Examples of such datasets include genomic databases, time-series signals, and neural network embeddings.
Large language models such as GPT~\cite{2020arXiv200514165B, OpenAI2023GPT4TR} and LLAMA-2~\cite{Touvron2023Llama2O}, and image embedding models~\cite{radford2021learning, dosovitskiy2020image} are a common source of neural network embeddings.
Among biological datasets,  SILVA 18S~\cite{10.1093/nar/gks1219} includes ribosomal DNA sequences from approximately 2.25 million genomes, reaching 50,000 letters when aligned.
Among time-series datasets, the RadioML dataset~\cite{oshea2018radioml} contains approximately 2.55 million samples of synthetically-generated radio frequency signals.

Similarity search is a fundamental task on such datasets, enabling applications such as recommendation~\cite{annoy} and classification ~\cite{suyanto2022knnclassifier}.
Yet as datasets grow in size and dimensionality, efficient and accurate similarity search becomes increasingly challenging;
even state-of-the-art algorithms exhibit a steep tradeoff between recall and throughput~\cite{malkov2016hnsw, johnson2019billion, annoy, aumuller2020ann}.

Given some measure of similarity between data points, two standard similarity search paradigms are $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN).
The former retrieves the $k$ most similar points to a query, while the latter retrieves all points within a similarity threshold $\rho$ of a query.
While prior work has used the term \textit{approximate} search to refer to $\rho$-NN, we use it more narrowly: an {approximate} search algorithm is any algorithm that yields less than perfect recall, whereas an \textit{exact} algorithm guarantees perfect recall.

$k$-NN search is one of the most widely-used methods in classification and recommendation~\cite{fix1952discriminatory, cover1967nearest}, but na\"{i}ve implementations are prohibitively slow because their time complexity is linear in the dataset's cardinality.
Existing fast algorithms are often approximate~\cite{gao2023high}, which may suffice for some applications, but the need for efficient and \textit{exact} search remains~\cite{ukey2023survey}.
For example, for a majority voting classifier, approximate $k$-NN search may agree with exact $k$-NN search for large values of $k$, but may be sensitive to local perturbations for smaller values of $k$.

For example, approximate methods may fail in majority-voting classifiers with small $k$ or poorly separated classes~\cite{zhang2022imbalanced}, and distance functions like cosine similarity, which violate the triangle inequality, have shown poor performance in biomedical applications~\cite{hu2016distance}.

This paper introduces CAKES (CLAM-Accelerated $K$-NN Entropy-Scaling Search), a set of three novel algorithms for \emph{exact} $k$-NN search.
We benchmark CAKES against state-of-the-art methods--FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy}--on datasets from the ANN-benchmarks suite~\cite{aumuller2020ann}.
We also evaluate performance on a large genomic dataset, the SILVA 18S dataset~\cite{10.1093/nar/gks1219} using Levenshtein~\cite{levenshtein1966binary} distance on unaligned genomic sequences, and on a radio frequency dataset, RadioML~\cite{oshea2018radioml}, using Dynamic Time Warping (DTW)~\cite{gold2018dynamic} distance on complex-valued time-series.
To further contextualize results, we compare against a synthetic dataset with simple statistical properties.

@@

Several algorithms have been developed to scale $k$-nearest neighbor search to large datasets, including Hierarchical Navigable Small World networks (HNSW)~\cite{malkov2016hnsw}, InVerted File indexing (FAISS-IVF)~\cite{faissivf}, random projection and tree building (ANNOY)~\cite{annoy}, and entropy-scaling search~\cite{yu2015entropy, ishaq2019clustered}. However, many of these algorithms do not support \emph{exact} search (as defined in Section \ref{sec:introduction}).

Hierarchical Navigable Small World networks~\cite{malkov2016hnsw} rely on navigable small world (NSW) networksand skip lists.
HNSW builds a multi-layered small-world graph~\cite{kleinberg2000navigation, boguna2009navigability} of the dataset, with queries and elements placed at each layer according to an exponentially-decaying probability distribution.
Each node connects to its $M$ nearest neighbors (a tunable parameter) in the in the graph, where $M$. Search proceeds top-down, using a greedy search at each layer. Accuracy is controlled by the $efSearch$ hyperparameter, which specifies how many candidates are explored during search.

InVerted File indexing (IVF)~\cite{faissivf, sacks1987multikey, kent1990signature} partitions data into high-dimensional Voronoi cells, and searches exhaustively only within cells near the query.
The number of clusters is set via the $n_{list}$ parameter; accuracy is improved by probing additional neighboring cells, controlled by the $n_{probe}$ parameter. This approach resembles earlier entropy-scaling methods~\cite{yu2015entropy}.

ANNOY~\cite{annoy} uses random projection and tree building. 
At each intermediate node of the tree, a random hyperplane, defined by two sampled points, splits the space.  Multiple such trees form a forest; increasing the number of trees improves recall but increases search time.

@@

Entropy-scaling algorithms exploit the inherent structure of large datasets, achieving complexity that scales with topological properties—such as metric entropy and fractal dimension—rather than cardinality. This makes them especially suitable for manifold-structured data.

In 2019, we introduced CHESS (Clustered Hierarchical Entropy-Scaling Search)~\cite{ishaq2019clustered}, which extended the original flat clustering approach to entropy-scaling $\rho$-NN search from a flat clustering approach~\cite{yu2015entropy} to a hierarchical one.
CLAM (Clustering, Learning and Approximation with Manifolds), originally developed to allow ``manifold mapping'' for anomaly detection~\cite{ishaq2021clustered}, refined the clustering algorithm from CHESS.

This paper introduces CAKES, a set of three entropy-scaling algorithms for $k$-NN search, implemented in Rust. We also provide a theoretical analysis of their time complexity in Sections~\ref{sec:methods:knn-search:repeated-rnn-complexity} and~\ref{sec:methods:knn-search:complexity-of-sieve-methods}.
These analyses are not worst-case analyses in the traditional sense, as they do not assume the worst possible dataset, namely, a uniform distribution.
Given that CAKES's algorithms are intended to be used on datasets with a manifold structure, complexity analysis assuming a uniform distribution of data would be uninformative.
As a result, our analysis assumes datasets with manifold structure—characterized by low fractal dimension and metric entropy—reflecting the conditions for which CAKES is designed.
